Gravitational-wave sensitivity curves

### 2014-07-07 Sources.tex
addressed some of Christopher's comments

All the sources described here are represented by shaded boxes in \ref{app:a}. The boxes are drawn in such a way that there is a reasonable event rate for sources at a detectable SNR. Descriptions of each source are given in the following sections. Sources with short durations (i.e., burst sources) and sources which evolve in time over much longer timescales than our observations are drawn with flat-topped boxes \cplb{for $h_\mathrm{c}(f)$}. Inspiralling binaries, or stochastic backgrounds of binaries, are drawn with a sloping top proportional to $f^{-2/3}$ \cplb{for $h_\mathrm{c}(f)$}, which is result derived in section \ref{sec:stoch}. The width of the box gives the range of frequencies sources of a given type can have a while remaining at a detectable amplitude.

@@

Space-based detectors shall be sensitive to equal-mass mergers in the range $(10^{4}$--$10^{7})\,\Msun$. Predictions of the event rate for these mergers range from ${\mathcal{O}}(3$--$300)~\mathrm{yr}^{-1}$ for LISA with SNRs of up to $10^4$ \cplb{[citation needed]}. The large range in the rate reflects our uncertainty in the growth mechanisms of the supermassive black hole population \citep{Volonteri2010}. Plotted in \ref{app:a} are boxes with a ratio $h_\mathrm{c}/h_{n}=100$ for eLISA at its peak sensitivity. The range of frequencies plotted is $(3\times 10^{-4}$--$3\times 10^{-1})~\mathrm{Hz}$, this corresponds to circular binaries in the mass range quoted above.

@@

Galactic binaries divide into two classes: the unresolvable and the resolvable galactic binaries. The unresolvable binaries overlap to form a stochastic background as discussed in section \ref{sec:stoch}. \cplb{The distinction between resolvable and unresolvable is detector specific; here we choose LISA. This boundary will not be too different for eLISA but would move substantially for either of the decihertz detectors.} Plotted in \ref{app:a} is the estimate of this background due to \citet{Nelemans} where an observation time of one year has been assumed,
\begin{equation}
h_\mathrm{c}(f)= 5\times 10^{-21} \left(\frac{f}{10^{-3}~\mathrm{Hz}}\right)^{-2/3} \; .
 \end{equation}
Estimates for the event rate of resolvable binaries centre around ${\mathcal{O}}(10^{3})$ events for eLISA. The boxes plotted in \ref{app:a} have a ratio $h_\mathrm{c}/h_{n}=50$ for eLISA at its peak sensitivity. The frequency range of the box is $\left(3\times10^{-4}\right.$--$\left.10^{-2}\right)~\mathrm{Hz}$, estimated from Monte Carlo population simulation results presented in \citet{Amaro-Seoane-et-al}.

@@

EMRIs occur when a compact stellar mass object inspirals into a supermassive black hole. There is extreme uncertainty in the event rate for EMRIs due to the poorly constrained astrophysics in galactic centres \citep[e.g.,][]{Merritt2011}; the best guess estimate is ${\mathcal{O}}(10)$ events per year with eLISA with SNR $\ge 20$ \cplb{[citation needed]}. The boxes plotted in \ref{app:a} have a characteristic strain of $h_{c}=3\times 10^{-20}$ at $10^{-2}~\mathrm{Hz}$, which corresponds to a $10\Msun$ BH inspiralling into a $10^{6}\Msun$ black hole at a luminosity distance of $1~\mathrm{Gpc}$. The frequency width of the box is somewhat unknown; EMRI events can occur into a black hole of any mass, and hence EMRIs can occur at any frequency. However, there is substantial uncertainty in the black hole mass function \cplb{[citation needed]}, so the boxes in \ref{app:a} are drawn with a width comparable to that of the LISA sensitivity curve.

>>>>

All the sources described here are represented by shaded boxes in \ref{app:a}. The boxes are drawn in such a way that there is a reasonable event rate for sources at a detectable SNR. Descriptions of each source are given in the following sections. Sources with short durations (i.e., burst sources) and sources which evolve in time over much longer timescales than our observations are drawn with flat-topped boxes for $h_\mathrm{c}(f)$. Inspiralling binaries, or stochastic backgrounds of binaries, are drawn with a sloping top proportional to $f^{-2/3}$ for $h_\mathrm{c}(f)$, which is result derived in section \ref{sec:stoch}. The width of the box gives the range of frequencies sources of a given type can have a while remaining at a detectable amplitude.

@@

Space-based detectors shall be sensitive to equal-mass mergers in the range $(10^{4}$--$10^{7})\,\Msun$. Predictions of the event rate for these mergers range from ${\mathcal{O}}(10$--$100)~\mathrm{yr}^{-1}$ for eLISA with SNRs of up to $10^3$ \citep{TheGravitationalUniverse}. The large range in the rate reflects our uncertainty in the growth mechanisms of the supermassive black hole population \citep{Volonteri2010}. Plotted in \ref{app:a} are boxes with a ratio $h_\mathrm{c}/h_{n}=100$ for eLISA at its peak sensitivity. The range of frequencies plotted is $(3\times 10^{-4}$--$3\times 10^{-1})~\mathrm{Hz}$, this corresponds to circular binaries in the mass range quoted above.

@@

Galactic binaries divide into two classes: the unresolvable and the resolvable galactic binaries. The unresolvable binaries overlap to form a stochastic background as discussed in section \ref{sec:stoch}. The distinction between resolvable and unresolvable is detector specific; here we choose LISA. This boundary will not be too different for eLISA but would move substantially for either of the decihertz detectors. Plotted in \ref{app:a} is the estimate of this background due to \citet{Nelemans} where an observation time of one year has been assumed,
\begin{equation}
h_\mathrm{c}(f)= 5\times 10^{-21} \left(\frac{f}{10^{-3}~\mathrm{Hz}}\right)^{-2/3} \; .
 \end{equation}
Estimates for the event rate of resolvable binaries centre around ${\mathcal{O}}(10^{3})$ events for eLISA. The boxes plotted in \ref{app:a} have a ratio $h_\mathrm{c}/h_{n}=50$ for eLISA at its peak sensitivity. The frequency range of the box is $\left(3\times10^{-4}\right.$--$\left.10^{-2}\right)~\mathrm{Hz}$, estimated from Monte Carlo population simulation results presented in \citet{Amaro-Seoane-et-al}.

@@

EMRIs occur when a compact stellar mass object inspirals into a supermassive black hole. There is extreme uncertainty in the event rate for EMRIs due to the poorly constrained astrophysics in galactic centres \citep[e.g.,][]{Merritt2011}; the best guess estimate is around $25$ events per year with eLISA with SNR $\ge 20$ \citep{TheGravitationalUniverse}. The boxes plotted in \ref{app:a} have a characteristic strain of $h_{c}=3\times 10^{-20}$ at $10^{-2}~\mathrm{Hz}$, which corresponds to a $10\Msun$ BH inspiralling into a $10^{6}\Msun$ black hole at a luminosity distance of $1~\mathrm{Gpc}$. The frequency width of the box is somewhat unknown; EMRI events can occur into a black hole of any mass, and hence EMRIs can, in principal, occur at any frequency, so the boxes in \ref{app:a} are drawn with a width comparable to that of the LISA sensitivity curve.

### 2014-07-07 Typesofsource.tex
addressed some of Christopher's comments

\subsection{Inspirals}\label{sec:insp}
Inspiralling binaries spend a variable amount of time in each frequency band. If $\phi$ is the orbital phase then the number of cycles generated at frequency $f$ can be estimated as
\begin{equation}\label{eq:inspiral} {\cal{N}}_{\textrm{cycles}} = \frac{f}{2\pi}\frac{\textrm{d}\phi}{\textrm{d}f}=\frac{f^{2}}{\dot{f}} \; ,\end{equation}
where an overdot represents the time derivative and $f=\dot{\phi}$. The squared SNR will scale with ${\cal{N}}_{\textrm{cycles}}$, so it would be expected that $h_{c}(f)\approx \sqrt{{\cal{N}}_{\textrm{cycles}}}\tilde{h}(f)$. The exact conversion can be derived by considering the Fourier transform in the stationary phase approximation. Consider a source signal with approximately constant amplitude, $h$, and central frequency $f'$.
\begin{eqnarray}&h(t)= h\exp\left(\rmi \phi(t)\right) \\
\Rightarrow &\tilde{h}(f)=\int_{-\infty}^{\infty}d\textrm{t}\;h\exp \left(2\pi \rmi \left(\frac{\phi(t)}{2\pi t}-f\right)t\right)\end{eqnarray}
Without loss of generality assume an initial phase of zero and expand the exponent in powers of $t$,
\begin{eqnarray} \label{eq:FTofEMRI} \tilde{h}(f) &\approx \int_{-\infty}^{\infty}\rmd t\; h \exp\left[2\pi\rmi \dot{f'} \left(t^{2}+\frac{f'-f}{\dot{f'}}t\right)\right]\nonumber \\
&\approx h\exp\left[-\pi\rmi\frac{\left(f'-f\right)^{2}}{2\dot{f'}}\right]\int_{-\infty}^{\infty}\rmd t\;\exp\left[2\pi\rmi\dot{f'}\left(t+\frac{f'-f}{2\dot{f'}}\right)^{2}\right] \nonumber \\
&\approx\frac{h}{\sqrt{2i\dot{f'}}}\exp\left[-\pi\rmi\frac{\left(f'-f\right)^{2}}{2\dot{f'}}\right]\; .
\end{eqnarray}
From (\ref{eq:strains}) and (\ref{eq:FTofEMRI}) the characteristic strain and the Fourier transform can be related for inspiralling sources \citep{FinnThorne}
\begin{equation}\label{eq:insphc}h_{c}(f) = \sqrt{\frac{2f^{2}}{\dot{f}}}h\;.\end{equation}
Equation (\ref{eq:insphc}) is the relation between $h_{c}(f)$ and the instantaneous amplitude $h$ for an inspiralling source; for other types of source a new relation which satisfies (\ref{eq:hc}) will have to be found. Equation (\ref{eq:hc}) should be considered as the definition of characteristic strain and (\ref{eq:insphc}) a consequence of it for inspirals.

>>>>

\subsection{Inspirals}\label{sec:insp}
Inspiralling binaries spend a variable amount of time in each frequency band. If $\phi$ is the orbital phase then the number of cycles generated at frequency $f$ can be estimated as
\begin{equation}\label{eq:inspiral} {\cal{N}}_{\textrm{cycles}} = \frac{f}{2\pi}\frac{\textrm{d}\phi}{\textrm{d}f}=\frac{f^{2}}{\dot{f}} \; ,\end{equation}
where an overdot represents the time derivative and $f=\dot{\phi}$. The squared SNR will scale with ${\cal{N}}_{\textrm{cycles}}$, so it would be expected that $h_{c}(f)\approx \sqrt{{\cal{N}}_{\textrm{cycles}}}\tilde{h}(f)$. The exact conversion can be derived by considering the Fourier transform in the stationary phase approximation. Consider a source signal with approximately constant amplitude, $h$, and central frequency $f'$.
\begin{eqnarray}&h(t)= h\exp\left(\rmi \phi(t)\right) \\
\Rightarrow &\tilde{h}(f)=\int_{-\infty}^{\infty}d\textrm{t}\;h\exp \left(2\pi \rmi \left(\frac{\phi(t)}{2\pi t}-f\right)t\right)\end{eqnarray}
Without loss of generality assume an initial phase of zero and expand the exponent in powers of $t$,
\begin{eqnarray} \label{eq:FTofEMRI} \tilde{h}(f) &\approx \int_{-\infty}^{\infty}\rmd t\; h \exp\left[2\pi\rmi \dot{f'} \left(t^{2}+\frac{f'-f}{\dot{f'}}t\right)\right]\nonumber \\
&\approx h\exp\left[-\pi\rmi\frac{\left(f'-f\right)^{2}}{2\dot{f'}}\right]\int_{-\infty}^{\infty}\rmd t\;\exp\left[2\pi\rmi\dot{f'}\left(t+\frac{f'-f}{2\dot{f'}}\right)^{2}\right] \nonumber \\
&\approx\frac{h}{\sqrt{2i\dot{f'}}}\exp\left[-\pi\rmi\frac{\left(f'-f\right)^{2}}{2\dot{f'}}\right]\; .
\end{eqnarray}
From (\ref{eq:differentdescriptions}) and (\ref{eq:FTofEMRI}) the characteristic strain and the Fourier transform can be related for inspiralling sources \citep{FinnThorne}
\begin{equation}\label{eq:insphc}h_{c}(f) = \sqrt{\frac{2f^{2}}{\dot{f}}}h\;.\end{equation}
Equation (\ref{eq:insphc}) is the relation between $h_{c}(f)$ and the instantaneous amplitude $h$ for an inspiralling source; for other types of source a new relation which satisfies (\ref{eq:hc}) will have to be found. Equation (\ref{eq:hc}) should be considered as the definition of characteristic strain and (\ref{eq:insphc}) a consequence of it for inspirals.
